Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-06-03 ms/spaceness.tex
biorxiv ms

\documentclass[11pt,twoside,lineno]{preprint}
\usepackage{amsmath}
\usepackage[nodisplayskipstretch]{setspace}
\usepackage{makecell} 
\usepackage{changepage}
\usepackage{afterpage}
\setstretch{1.5}
\NeedsTeXFormat{LaTeX2e}
\newcommand*{\articletypename}{PREPRINT}
% Use the documentclass option 'lineno' to view line numbers

@@

\textcolor{red}{Two areas in which spatial structure may be particularly important are demographic inference and genome-wide association studies (GWAS). Previous work has found that discrete population structure can create false signatures of population bottlenecks when attempting to infer demographic histories from microsatellite variation \citep{Chikhi2010}, statistics summarizing the site frequency spectrum (SFS) \citep{Ptak2002,Stadler2009,Onge2012}, or runs of homozygosity in a single individual \citep{Mazet2015}. The increasing availability of whole-genome data has led to the development of many methods that attempt to infer detailed trajectories of population sizes through time based on a variety of summaries of genetic data \citep{Liu2015,Schiffels2014,Sheehan2013,Terhorst2016}. 
Because all of these methods assume that the populations being modeled are approximately randomly mating, they are likely affected by spatial biases in the genealogy of sampled individuals \citep{Wakeley1999}, which may lead to incorrect inference of population changes over time \citep{Mazet2015}. However, previous investigations of these effects have focused on discrete rather than continuous space models, and the level of isolation by distance at which inference of population size trajectories become biased by structure is not well known. Here we test how two methods suitable for use with large samples of individuals -- stairwayplot \citep{Liu2015} and SMC++ \citep{Terhorst2016} -- perform when applied to populations evolving in continuous space with varying sampling strategies and levels of dispersal. }

@@

To study this, we have implemented an individual-based model in continuous geography 
that incorporates overlapping generations, local dispersal of offspring, and density-dependent survival. 
We simulate chromosome-scale genomic data in tens of thousands of individuals from parameter regimes relevant to common subjects of population genetic investigation such as humans and Drosophila, and output the full genealogy and recombination history of all final-generation individuals. 
We use these simulations to test how sampling strategy interacts with geographic population structure to cause systematic variation in population genetic summary statistics 
typically analyzed assuming discrete population models. 
We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum. 
Last, we examine the impacts of continuous geography on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.

@@

% The best-studied approaches to population genetics in continuous space were developed by \citet{Wright1942,Wright1943} and \citet{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. 
The degree to which genetic relationships are geographically correlated
depends on the chance that two geographically nearby individuals are close relatives
-- in modern terms, by the tension between
migration (the chance that one is descended from a distant location)
and coalescence (the chance that they share a parent).
A key early observation \citep[by][]{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
or the number of potential parents of a randomly chosen individual. Empirical estimates of neighborhood size vary hugely across species
-- even in human populations,
estimates range from 40 to over 5,000 depending on the population and method of estimation (Table~\ref{table:NStable}).
%\citet{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population approaches the random mating expectation when $\rho\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

The first approach to modeling continuously-distributed populations
was to endow individuals in a Wright-Fisher model with locations in continuous space.
However, since the total size of the population is constrained, this introduces interactions between arbitrarily distant individuals,
which (aside from being implausible) 
was shown by \citet{Felsenstein1975} to eventually lead to unrealistic population clumping
if the range is sufficiently large.
Another method for modeling spatial populations is to assume the existence of a grid of discrete randomly-mating populations connected by migration, 
thus enforcing regular population density by edict.
Among many other important results drawn from this class of ``lattice'' or ``stepping stone'' models, 
\citet{Rousset1997} showed that the slope of the linear regression of genetic differentiation ($F_{ST}$) against the logarithm of spatial distance is an estimate of neighborhood size. 
Although these grid models may be good approximations of continuous geography in many situations,
they do not model demographic fluctuations, and limit investigation of spatial structure below the level of the deme,
assumptions whose impacts are unknown. 
An alternative method for dealing with continuous geography is a new class of coalescent models,
the Spatial Lambda Fleming-Viot models \citep{Barton2010,Kelleher2014}.

To avoid questionable assumptions, we here used forwards-time, individual-based simulations. By scaling the probability of survival in each timestep to local population density we shift reproductive output towards low-density regions, which prevents populations from clustering. Such models have been used extensively in ecological modeling but rarely in population genetics, where to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ where $L$ is the expected lifespan.
Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
in both the $x$ and $y$ coordinates, reflected to stay within the species range.
Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2 / 2 \sigma^2)$.

@@

%%%%%%%%%%%%%
\subsection{Summary Statistics}
We calculated the site frequency spectrum and a set of 18 summary statistics (Table \ref{table:sumstats}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. 
Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{IS}$), and Tajima's $D$, 
as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distance \citep{Rousset1997}, 
which we summarized as the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals.

@@

To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear regression GWAS with principal components as covariates in PLINK \citep{PURCELL2007}. SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis. Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe \ak{need citation}. Conceptually our approach is similar to that taken by \citet{Mathieson2012}, though here we model fully continuous spatial variation and compare GWAS output across a range of dispersal distances.

@@

Generation time similarly shows complex behavior with respect to neighborhood sizes, and varies between 5.2 and 4.9 timesteps per genration across the parameter range explored. 
% This occurs because individuals may reproduce more than once over their lifespan, and the lifespan of an individual is determined by density-dependent competition the model. 
Under both the spatial and random mating models, 
generation time reaches a minimum at a neighborhood size of around 50. 
Interestingly, under the range of neighborhood sizes that we examined, generation times between the random mating and spatial models are never quite equivalent -- presumably this would cease to be the case at neighborhood sizes higher than we simulated here.

@@

Even though certain aspects of population demography depend on the scale of spatial interactions, it still could be that population genetic variation is well-described by a well-mixed population model. Indeed, mathematical results suggest that genetic variation in some spatial models should be well-approximated by a Wright-Fisher population if neighborhood size is large and all samples are geographically widely separated \citep{wilkins2004separation,zahle2005stepping}. However, the behavior of most common population genetic summary statistics \textcolor{red}{other than Tajima's $D$ \citep{Stadler2009}} has not yet been described in realistic geographic models. Moreover, as we will show, spatial sampling strategies can affect summaries of variation at least as strongly as the underlying population dynamics.

@@

Figure \ref{fig:sumstats} shows the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure~\ref{fig:sumstats}A) and several standard population genetic summary statistics (Figure~\ref{fig:sumstats}B). \textcolor{red}{Consistent with findings in island and stepping stone simulations \citep{Stadler2009}, the SFS shows a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation. This bias is most pronounced below neighborhood sizes $\leq 100$ and is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}).} 
Reflecting this, Tajima's $D$ is quite positive in the same situations (Figure~\ref{fig:sumstats}B). Notably, the point at which Tajima's $D$ approaches 0 differs strongly across sampling strategies -- varying from a neighborhood size of roughly 50 for random sampling to at least 1000 for midpoint sampling.

One of the most commonly used summaries of variation is Tajima's summary of nucleotide divergence, $\theta_{\pi}$, calculated as the mean density of nucleotide differences averaged across pairs of samples. As can be seen in Figure \ref{fig:sumstats}B, $\theta_{\pi}$ differs as much as nearly three-fold between the random mating and spatial models. $\theta_{\pi}$ using each sampling strategy approaches the random mating expectation at its own rate, but by a neighborhood size of around 100 all models are equivalent. The differences between spatial and random mating simulations are much greater than expected from differences in census size divided by variance in offspring number. This likely occurs because $\theta_{\pi}$ is a measure of mean time to most recent common ancestor between two samples, and at small values of $\sigma$, the time for dispersal to mix ancestry across the range exceeds the mean nonspatial coalescent time. (For instance, at the smallest value of $\sigma=0.2$, the range is 250 dispersal distances wide, and since the location of a diffusively moving lineage after $k$ generations has variance $k \sigma^2$, it takes around $250^2 = 62500$ generations to mix across the range, which is roughly ten times larger than the random mating effective population size.) Interestingly, the effect of sampling strategy is reversed relative to that observed in Tajima's D -- midpoint sampling reaches random mating expectations around neighborhood size 50, while random sampling is inflated until around neighborhood size 100.

Values of observed heterozygosity and its derivative $F_{IS}$ also depend heavily on neighborhood size under spatial models as well as the sampling scheme. $F_{IS}$ is inflated above the expectation across most of the parameter space examined and across all sampling strategies. This effect is caused by a deficit of heterozygous individuals in low-dispersal simulations -- a continuous-space version of the Wahlund effect \citep{Wahlund1928}. Indeed, for random sampling under the spatial model, $F_{IS}$ does not approach the random mating equivalent until neighborhood sizes of nearly 1000. On the other hand, the dependency of raw observed heterozygosity on neighborhood size is not monotone. Under midpoint sampling observed heterozygosity is inflated even over the random mating expectation, as a result of the a higher proportion of heterozygotes occurring in the middle of the landscape (Figure~\ref{fig:hetmap}). This squares with a report from \citet{Shirk2014} who observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.

@@

\plr{median of how many? should be fig S4 here instead?}
Figure \ref{fig:demography}A shows the median inferred population size histories from each method, grouped by neighborhood size and sampling strategy. In general these methods tend to slightly overestimate ancient population sizes and infer recent population declines when neighborhood sizes are below 20 and sampling is spatially clustered (Figure \ref{fig:demography}A, Figure \ref{fig:demography_supplement}). The overestimation of ancient population sizes however is relatively minor, averaging around a two-fold inflation at 10,000 generations before present in the worst-affected bins. For stairwayplot we found that many runs infer dramatic population bottlenecks in the last 1,000 generations when sampling is spatially concentrated, resulting in ten-fold or greater underestimates of recent population sizes. However SMC++ appeared more robust to this error, with runs on point- and midpoint-sampled simulations at the lowest neighborhood sizes underestimating recent population sizes by roughly half and those on randomly-sampled simulations showing little error. Above neighborhood sizes of around 100, both methods performed relatively well when averaging across results from multiple simulations.

@@

The confounding effects of geographic structure are well known, 
and it is common practice to control for this by including principal components (PCs) as covariates to control for these effects.
This mostly works in our simulations -- after doing this, the vast majority of SNPs no longer surpass a significance threshold chosen to have a 5\% false discovery rate (FDR).
However, a substantial number of SNPs -- up to 1.5\% at the lowest dispersal distances -- still surpass this threshold (and thus would be false positives in a GWAS),
especially under ``corner'' and ``patchy'' environmental distributions (Figure~\ref{fig:gwas}C). 
At neighborhood sizes larger than 500, up to 0.31\% of SNPs were significant for corner and clinal environments. 
Given an average of 132,000 SNPs across simulations after MAF filtering, this translates to up to 382 false-positive associations; for human-sized genomes, this number would be much larger.
In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction (see example Manhattan plots in figure \ref{fig:manhattan_plots}).

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/gwas_summary_loglog.pdf}
    \caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS. Simulated quantitative phenotypes are determined only by an individual's location and the spatial distribution of environmental factors. In \textbf{A} we show the phenotypes and locations of sampled individuals under four environmental distributions, with transparency scaled to phenotype. As neighborhood size increases a PCA explains less of the total variation in the data (\textbf{B}). Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding PC positions as covariates (\textbf{C}). Quantile-quantile plots in \textbf{D} show inflation of $-\log_{10}(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
}
\label{fig:gwas}
\end{figure*}

@@

One of the most important differences between random mating and spatial models is the effect of sampling: 
in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation (Table \ref{table:sumstats}) , 
but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses \textcolor{red}{(see also \citet{Stadler2009})}. 
In most species,
the difficulty of traveling through all parts of a species range and the inefficiency of collecting single individuals at each sampling site 
means that most studies follow something closest to the ``point'' sampling strategy we simulated, 
in which multiple individuals are sampled from nearby points on the landscape. 
For example, in ornithology a sample of 10 individuals per species per locality is a common target when collecting for natural history museums. 
In classical studies of \textit{Drosophila} variation the situation is considerably worse, in which a single orchard might be extensively sampled.

@@

\textcolor{red}{Previous studies have found that population structure and nonrandom sampling can create spurious signals of population bottlenecks when attempting to infer demographic history with microsatellite variation, summary statistics, or runs of homozygosity \citep{Chikhi2010,Stadler2009,Ptak2002,Mazet2015}. Here we found that methods that infer detailed population trajectories through time based on the SFS and patterns of LD across the genome are also subject to this bias, with some combinations of dispersal and sampling strategy systematically inferring deep recent population bottlenecks and overestimating ancient $N_{e}$ by a around a factor of 2. We were surprised to see that both stairwayplot and SMC++ can tolerate relatively strong isolation by distance -- i.e., neighborhood sizes of 20 -- and still perform well when averaging results across multiple simulations. Inference in populations with neighborhood sizes over 20 was relatively unbiased unless samples were concentrated in the middle of the range (Figure \ref{fig:demography}). 
Although median demography estimates across many independent simulations were fairly accurate,
empirical work has only a single estimate to work with,
and individual model fits (Figure \ref{fig:demography_supplement})
suggest that spuriously inferred population size changes and bottlenecks are common, especially at small neighborhood sizes.
As we will discuss below, most empirical estimates of neighborhood size, including all estimates for human populations, are large enough that population size trajectories inferred by these approaches should not be strongly affected by spatial biases created by dispersal in continuous landscapes. In contrast, \citet{Mazet2015} found that varying migration rates through time could create strong biases in inferred population trajectories from an $n$-island model with parameters relevant for human history, suggesting that changes in migration rates through time are more likely to drive variation in inferred $N_{e}$ than isolation by distance.}

\textcolor{red}{We found that SMC++ was more robust to the effects of space than stairwayplot, underestimating recent populations by roughly half in the worst time periods rather than nearly 10-fold as with stairwayplot.} Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms. A more worrying pattern was the high level of variance in inferred $N_{e}$ trajectories for individual model fits using these methods, which was highest in simulations with the smallest neighborhood size (Figure \ref{fig:demography}, Figure \ref{fig:demography_supplement}). This suggests that, at a minimum, researchers working with empirical data should replicate analyses multiple times and take a rolling average if model fits are inconsistent across runs. Splitting samples and running replicates on separate subsets -- the closest an empirical study can come to our design of averaging the results from multiple simulations -- may also alleviate this issue.

@@

Though our simulation allows incorporation of realistic demographic and spatial processes, 
it is inevitably limited by the computational burden of tracking tens or hundreds of thousands of individuals in every generation. 
In particular, computations required for mate selection and spatial competition scale approximately with the product of the total census size and the neighborhood size and so increases rapidly for large populations and dispersal distances. 
The reverse-time model of continuous space evolution described by \citet{Barton2010} and implemented by \citet{Kelleher2014} 
allows exploration of parameter regimes with population and landscape sizes more directly comparable to empirical cases like humans. 
Alternatively, implementation of parallelized calculations may allow progress with forwards-time simulations.

>>>>

\documentclass[11pt,twoside,lineno,hidelinks]{preprint}
\usepackage{amsmath}
\usepackage[nodisplayskipstretch]{setspace}
\usepackage{makecell} 
\usepackage{changepage}
\usepackage{afterpage}
\setstretch{1.5}
\NeedsTeXFormat{LaTeX2e}
\newcommand*{\articletypename}{PREPRINT}
% Use the documentclass option 'lineno' to view line numbers

@@

Two areas in which spatial structure may be particularly important are demographic inference and genome-wide association studies (GWAS). Previous work has found that discrete population structure can create false signatures of population bottlenecks when attempting to infer demographic histories from microsatellite variation \citep{Chikhi2010}, statistics summarizing the site frequency spectrum (SFS) \citep{Ptak2002,Stadler2009,Onge2012}, or runs of homozygosity in a single individual \citep{Mazet2015}. The increasing availability of whole-genome data has led to the development of many methods that attempt to infer detailed trajectories of population sizes through time based on a variety of summaries of genetic data \citep{Liu2015,Schiffels2014,Sheehan2013,Terhorst2016}. 
Because all of these methods assume that the populations being modeled are approximately randomly mating, they are likely affected by spatial biases in the genealogy of sampled individuals \citep{Wakeley1999}, which may lead to incorrect inference of population changes over time \citep{Mazet2015}. However, previous investigations of these effects have focused on discrete rather than continuous space models, and the level of isolation by distance at which inference of population size trajectories become biased by structure is not well known. Here we test how two methods suitable for use with large samples of individuals -- stairwayplot \citep{Liu2015} and SMC++ \citep{Terhorst2016} -- perform when applied to populations evolving in continuous space with varying sampling strategies and levels of dispersal.

@@

To study this, we have implemented an individual-based model in continuous geography 
that incorporates overlapping generations, local dispersal of offspring, and density-dependent survival. 
We simulate chromosome-scale genomic data in tens of thousands of individuals from parameter regimes relevant to common subjects of population genetic investigation such as humans and \textit{Drosophila}, and output the full genealogy and recombination history of all final-generation individuals. 
We use these simulations to test how sampling strategy interacts with geographic population structure to cause systematic variation in population genetic summary statistics 
typically analyzed assuming discrete population models. 
We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum. 
Last, we examine the impacts of continuous geography on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.

@@

% The best-studied approaches to population genetics in continuous space were developed by \citet{Wright1942,Wright1943} and \citet{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. 
The degree to which genetic relationships are geographically correlated
depends on the chance that two geographically nearby individuals are close relatives
-- in modern terms, by the tension between
migration (the chance that one is descended from a distant location)
and coalescence (the chance that they share a parent).
A key early observation by Wright \citep{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
or the number of potential parents of a randomly chosen individual. Empirical estimates of neighborhood size vary hugely across species
-- even in human populations,
estimates range from 40 to over 5,000 depending on the population and method of estimation (Table~\ref{table:NStable}).
%\citet{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population approaches the random mating expectation when $\rho\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

The first approach to modeling continuously distributed populations
was to endow individuals in a Wright-Fisher model with locations in continuous space.
However, since the total size of the population is constrained, this introduces interactions between arbitrarily distant individuals,
which (aside from being implausible) 
was shown by \citet{Felsenstein1975} to eventually lead to unrealistic population clumping
if the range is sufficiently large.
Another method for modeling spatial populations is to assume the existence of a grid of discrete randomly mating populations connected by migration, 
thus enforcing regular population density by edict.
Among many other important results drawn from this class of ``lattice'' or ``stepping stone'' models, 
\citet{Rousset1997} showed that the slope of the linear regression of genetic differentiation ($F_{ST}$) against the logarithm of spatial distance is an estimate of neighborhood size. 
Although these grid models may be good approximations of continuous geography in many situations,
they do not model demographic fluctuations, and limit investigation of spatial structure below the level of the deme,
assumptions whose impacts are unknown. 
An alternative method for dealing with continuous geography is a new class of coalescent models,
the Spatial Lambda Fleming-Viot models \citep{Barton2010,Kelleher2014}.

To avoid questionable assumptions, we here used forward-time, individual-based simulations. By scaling the probability of survival in each timestep to local population density we shift reproductive output towards low-density regions, which prevents populations from clustering. Such models have been used extensively in ecological modeling but rarely in population genetics, where to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ where $L$ is the expected lifespan.
Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
in both the $x$ and $y$ coordinates.
Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2 / 2 \sigma^2)$.

@@

%%%%%%%%%%%%%
\subsection{Summary Statistics}
We calculated the site frequency spectrum and a set of 18 summary statistics (Table \ref{table:sumstats}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. 
Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{IS}$), and Tajima's $D$, 
as well as an isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distance analogous to \cite{Rousset1997}'s approach, 
which we summarized as the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals.

@@

To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear regression GWAS with principal components as covariates in PLINK \citep{PURCELL2007}. SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis. Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe \citep{Turchin2012,garcia2006,garcia2007}. Conceptually our approach is similar to that taken by \citet{Mathieson2012}, though here we model fully continuous spatial variation and compare GWAS output across a range of dispersal distances.

@@

Generation time similarly shows complex behavior with respect to neighborhood sizes, and varies between 5.2 and 4.9 timesteps per generation across the parameter range explored. 
% This occurs because individuals may reproduce more than once over their lifespan, and the lifespan of an individual is determined by density-dependent competition the model. 
Under both the spatial and random mating models, 
generation time reaches a minimum at a neighborhood size of around 50. 
Interestingly, under the range of neighborhood sizes that we examined, generation times between the random mating and spatial models are never quite equivalent -- presumably this would cease to be the case at neighborhood sizes higher than we simulated here.

@@

Even though certain aspects of population demography depend on the scale of spatial interactions, it still could be that population genetic variation is well-described by a well-mixed population model. Indeed, mathematical results suggest that genetic variation in some spatial models should be well-approximated by a Wright-Fisher population if neighborhood size is large and all samples are geographically widely separated \citep{wilkins2004separation,zahle2005stepping}. However, the behavior of most common population genetic summary statistics other than Tajima's $D$ \citep{Stadler2009} has not yet been described in realistic geographic models. Moreover, as we will show, spatial sampling strategies can affect summaries of variation at least as strongly as the underlying population dynamics.

@@

Figure \ref{fig:sumstats} shows the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure~\ref{fig:sumstats}A) and several standard population genetic summary statistics (Figure~\ref{fig:sumstats}B). Consistent with findings in island and stepping stone simulations \citep{Stadler2009}, the SFS shows a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation. This bias is most pronounced below neighborhood sizes $\leq 100$ and is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}). 
Reflecting this, Tajima's $D$ is quite positive in the same situations (Figure~\ref{fig:sumstats}B). Notably, the point at which Tajima's $D$ approaches 0 differs strongly across sampling strategies -- varying from a neighborhood size of roughly 50 for random sampling to at least 1000 for midpoint sampling.

One of the most commonly used summaries of variation is Tajima's summary of nucleotide divergence, $\theta_{\pi}$, calculated as the mean density of nucleotide differences averaged across pairs of samples. As can be seen in Figure \ref{fig:sumstats}B, $\theta_{\pi}$ in the spatial model is inflated by up to three-fold relative to the random mating model. This pattern is opposite the expectation from census population size (Figure \ref{fig:genparams}), because the spatial model has \emph{lower} census size than the random mating model at neighborhood sizes less than 100. Differences between these models likely occur because $\theta_{\pi}$ is a measure of mean time to most recent common ancestor between two samples, and at small values of $\sigma$, the time for dispersal to mix ancestry across the range exceeds the mean coalescent time under random mating. (For instance, at the smallest value of $\sigma=0.2$, the range is 250 dispersal distances wide, and since the location of a diffusively moving lineage after $k$ generations has variance $k \sigma^2$, it takes around $250^2 = 62500$ generations to mix across the range, which is roughly ten times larger than the random mating effective population size). $\theta_{\pi}$ using each sampling strategy approaches the random mating expectation at its own rate, but by a neighborhood size of around 100 all models are roughly equivalent. Interestingly, the effect of sampling strategy is reversed relative to that observed in Tajima's D -- midpoint sampling reaches random mating expectations around neighborhood size 50, while random sampling is inflated until around neighborhood size 100.

Values of observed heterozygosity and its derivative $F_{IS}$ also depend heavily on neighborhood size under spatial models as well as the sampling scheme. $F_{IS}$ is inflated above the expectation across most of the parameter space examined and across all sampling strategies. This effect is caused by a deficit of heterozygous individuals in low-dispersal simulations -- a continuous-space version of the Wahlund effect \citep{Wahlund1928}. Indeed, for random sampling under the spatial model, $F_{IS}$ does not approach the random mating equivalent until neighborhood sizes of nearly 1000. On the other hand, the dependency of raw observed heterozygosity on neighborhood size is not monotone. Under midpoint sampling observed heterozygosity is inflated even over the random mating expectation, as a result of the a higher proportion of heterozygotes occurring in the middle of the landscape (Figure~\ref{fig:hetmap}). This echoes a report from \citet{Shirk2014} who observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.

@@

Figure \ref{fig:demography}A shows the median inferred population size histories from each method across all simulations, grouped by neighborhood size and sampling strategy. In general these methods tend to slightly overestimate ancient population sizes and infer recent population declines when neighborhood sizes are below 20 and sampling is spatially clustered (Figure \ref{fig:demography}A, Figure \ref{fig:demography_supplement}). The overestimation of ancient population sizes however is relatively minor, averaging around a two-fold inflation at 10,000 generations before present in the worst-affected bins. For stairwayplot we found that many runs infer dramatic population bottlenecks in the last 1,000 generations when sampling is spatially concentrated, resulting in ten-fold or greater underestimates of recent population sizes. However SMC++ appeared more robust to this error, with runs on point- and midpoint-sampled simulations at the lowest neighborhood sizes underestimating recent population sizes by roughly half and those on randomly sampled simulations showing little error. Above neighborhood sizes of around 100, both methods performed relatively well when averaging across results from multiple simulations.

@@

The confounding effects of geographic structure are well known, 
and it is common practice to control for this by including principal components (PCs) as covariates to control for these effects.
This mostly works in our simulations -- after incorporating the first ten PC axes as covariates, the vast majority of SNPs no longer surpass a significance threshold chosen to have a 5\% false discovery rate (FDR).
However, a substantial number of SNPs -- up to 1.5\% at the lowest dispersal distances -- still surpass this threshold (and thus would be false positives in a GWAS),
especially under ``corner'' and ``patchy'' environmental distributions (Figure~\ref{fig:gwas}C). 
At neighborhood sizes larger than 500, up to 0.31\% of SNPs were significant for corner and clinal environments. 
Given an average of 132,000 SNPs across simulations after MAF filtering, this translates to up to 382 false-positive associations; for human-sized genomes, this number would be much larger.
In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction (see example Manhattan plots in figure \ref{fig:manhattan_plots}).

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/gwas_summary_loglog.pdf}
    \caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS. Simulated quantitative phenotypes are determined only by an individual's location and the spatial distribution of environmental factors. In \textbf{A} we show the phenotypes and locations of sampled individuals under four environmental distributions, with transparency scaled to phenotype. As neighborhood size increases a PCA explains less of the total variation in the data (\textbf{B}). Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding the first 10 PC coordinates as covariates (\textbf{C}). Quantile-quantile plots in \textbf{D} show inflation of $-\log_{10}(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
}
\label{fig:gwas}
\end{figure*}

@@

One of the most important differences between random mating and spatial models is the effect of sampling: 
in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation (Table \ref{table:sumstats}) , 
but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses (see also \citet{Stadler2009}). 
In most species,
the difficulty of traveling through all parts of a species range and the inefficiency of collecting single individuals at each sampling site 
means that most studies follow something closest to the ``point'' sampling strategy we simulated, 
in which multiple individuals are sampled from nearby points on the landscape. 
For example, in ornithology a sample of 10 individuals per species per locality is a common target when collecting for natural history museums. 
In classical studies of \textit{Drosophila} variation the situation is considerably worse, in which a single orchard might be extensively sampled.

@@

Previous studies have found that population structure and nonrandom sampling can create spurious signals of population bottlenecks when attempting to infer demographic history with microsatellite variation, summary statistics, or runs of homozygosity \citep{Chikhi2010,Stadler2009,Ptak2002,Mazet2015}. Here we found that methods that infer detailed population trajectories through time based on the SFS and patterns of LD across the genome are also subject to this bias, with some combinations of dispersal and sampling strategy systematically inferring deep recent population bottlenecks and overestimating ancient $N_{e}$ by a around a factor of 2. We were surprised to see that both stairwayplot and SMC++ can tolerate relatively strong isolation by distance -- i.e., neighborhood sizes of 20 -- and still perform well when averaging results across multiple simulations. Inference in populations with neighborhood sizes over 20 was relatively unbiased unless samples were concentrated in the middle of the range (Figure \ref{fig:demography}). 
Although median demography estimates across many independent simulations were fairly accurate,
empirical work has only a single estimate to work with,
and individual model fits (Figure \ref{fig:demography_supplement})
suggest that spuriously inferred population size changes and bottlenecks are common, especially at small neighborhood sizes.
As we will discuss below, most empirical estimates of neighborhood size, including all estimates for human populations, are large enough that population size trajectories inferred by these approaches should not be strongly affected by spatial biases created by dispersal in continuous landscapes. In contrast, \citet{Mazet2015} found that varying migration rates through time could create strong biases in inferred population trajectories from an $n$-island model with parameters relevant for human history, suggesting that changes in migration rates through time are more likely to drive variation in inferred $N_{e}$ than isolation by distance.

We found that SMC++ was more robust to the effects of space than stairwayplot, underestimating recent populations by roughly half in the worst time periods rather than nearly 10-fold as with stairwayplot. Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms. A more worrying pattern was the high level of variance in inferred $N_{e}$ trajectories for individual model fits using these methods, which was highest in simulations with the smallest neighborhood size (Figure \ref{fig:demography}, Figure \ref{fig:demography_supplement}). This suggests that, at a minimum, researchers working with empirical data should replicate analyses multiple times and take a rolling average if model fits are inconsistent across runs. Splitting samples and running replicates on separate subsets -- the closest an empirical study can come to our design of averaging the results from multiple simulations -- may also alleviate this issue.

@@

Though our simulation allows incorporation of realistic demographic and spatial processes, 
it is inevitably limited by the computational burden of tracking tens or hundreds of thousands of individuals in every generation. 
In particular, computations required for mate selection and spatial competition scale approximately with the product of the total census size and the neighborhood size and so increase rapidly for large populations and dispersal distances. 
The reverse-time model of continuous space evolution described by \citet{Barton2010} and implemented by \citet{Kelleher2014} 
allows exploration of parameter regimes with population and landscape sizes more directly comparable to empirical cases like humans. 
Alternatively, implementation of parallelized calculations may allow progress with forward-time simulations.
